Venn diagrams as indicators of erroneous statistical reasoning in transcriptomics.

### 2022-01-18 mk4/response_to_editor.md

Thanks for your responses to addressing the reviewer's concerns. I would
like to ask the authors to address the following concerns regarding the
paper. These are essential before accepting the paper.

The authors state that "Here we show that both, the use of Venn diagrams to
find genes which are thought to be specific for a certain comparison, as
well as gene set enrichment analysis applied to such subsets is a fallacy.
A statistically correct approach involves testing for interaction."

To start with, no one uses "Venn diagrams to find genes which are thought to be
specific for a certain comparison". It is just a visualization tool. 
The flaw
is not in the Venn diagram but in how one could interpret it. Also, applying
gene set enrichment analysis to such subsets could be done even without the use
of Venn diagrams.

The title and abstract are misleading and more journalistic
than scientific.  An analogous example to how the authors presented the paper
would be to condemn the use of a knife for murders. 
This style is misleading
and not acceptable for a scientific paper. The title and abstract should be
completely reformatted to focus on the underlying issue.

**We agree with these observations and understand your concerns. We have
reformulated abstract and discussion to state more clearly that Venn
diagrams are not the cause of flawed statistics but are very easily
misinterpreted and a readily identifiable symptom of erroneous statistical
reasoning, since we do observe a substantial fraction of papers using Venn
diagrams misinterpreting the results. Extending your metaphor, if knives
were like VDs in this context, then every fourth use of a knife would
result in a homicide.**

**We regret if our manuscript is perceived as too journalistic, even though
we think that communicating such a matter is to a large part an educational
mission and requires somewhat less complex and more pointed language. We
hope that our rewritten title, abstract and discussion strike a good
balance between scientific restraint and accessibility for
non-specialists.**

@@

The authors state that "We have noticed a widespread fallacy related to gene
set enrichment analysis" however, no single example (or reference) has been
provided.

**We do not want to single out specific papers or point fingers at specific
authors.  However, given that the sources of all our calculations are
available from github, we placed there the lists of links to papers which –
in our view – have incorrectly interpreted Venn diagrams. For example, the
file `literature_survey_sci_immunol.md` includes six (out of 14 analysed)
links to papers in Science Immunology which incorrectly interpreted Venn
diagrams. The file `literature_survey_scirep.md` contains 73 links to
papers from Scientific Reports (out of 238 reviewed). We now mention that
explicitely in "Methods".**

@@

The literature review results with incorrect use cases should be provided as
supplementary materials to support the claims made in the paper. 
Justification
for the claims made in the paper is essential.

**We have now included the following statement in the
Methods section:**

@@

*"The results of the literature survey (including links to papers classified
as incorrectly analysing the interaction) are included in the manuscript
sources."*

Authors state that "Of the 431 genes significant in G1, but not in G2, 199
(46%) are significant in the full data set; of the 278 genes significant in G2,
but not in G1, 99 (36%) are significant in the full data set. Given that G1 and
G2 were sampled from the total population, and since the FDR was set to 0.05,
we can assume that at least between a third and a half of the genes that
appeared to be specific" in the initial analysis were, in fact, false negatives
in one of the comparisons." The conclusion made is based on the assumption that
there are no false positives when analyzing the whole dataset. 
This is not
justifiable.

**Our conclusion is based on the assumption that the fraction of FP among
significant genes is controlled by the Benjamini-Hochberg procedure, in
other words, that the false discovery rate (FDR) is not more than 5% (as
stated in the quoted paragraph). This implies that in the whole data set
there are at most 30 DEGs which are FP. Since 199+99 = 298
supposedly specific genes are significant when analysing the whole data
set, even if all 30 FPs are among them, there are still 268 genes which are
not FP and which appear to be group specific. 100% * 268 / (431 + 278) =
37%. We have now made these simple calculations more explicit in the
text.**

Also, splitting samples of a dataset into two groups results in
smaller sample sizes for each group. 
This comes with two major shortcomings:
(1) the test for differential expression of genes are more sensitive to noise
and outliers, resulting in false positives
, and (2) the test for differential
expression of genes might not have the power to reject the null hypothesis (no
differential expression), resulting in introducing false negatives.

**We agree that splitting the dataset results in decreased statistical
power. However, we don't find it obvious that the false discovery rate will
increase, since it's controlled by the standard BH false discovery rate
procedure and the FDR cutoff of 0.05. Instead, we fully agree that FNs are
the core of the problem and used the FDR cutoff of 0.05 to not unduely
inflate the number of false negatives (a more conservative threshold would
result in a higher false negative rate). Even in most favorable
circumstances the FNR will likely be above 20%, especially for
transcriptomic studies (see also
https://academic.oup.com/bib/article/19/4/713/2920205).**

**In any case, there are 80 samples in our analysis (40 samples per
group and 20 per group/treatment combination), which is not a small sample
size for this type of transcriptomic analysis. In fact, a survey on RNA
sample size in real world data (Baccarella et al.) finds that half of the
studies in humans used sample sizes sizes of 6 or less per group. Nevertheless, we
welcome the suggestion to analyze the sample size dependence: in fact, the
larger the sample size, the more artefactual enrichments are observed,
because the differential expression yields more results and the
hypergeometric test gets more powerful with more DEGs. We have now included
a section comparing the results for 10, 20 (our example) and 40 samples per
group/treatment combination (thus, total sample sizes of 40, 80 and 160).
For each sample size, we have run 100 replicates of the group sampling,
replicated the whole procedure and gathered all test results. We hope that
the new results shown ion the new Fig. 4, C and D provide a useful
illustration of this reasoning.**

@@

Therefore, the design of the experiment is biased in favour of the
"desired" conclusion.

**We respectfully disagree. Our approach is quite similar to
cross-validation and a straightforward way to construct two groups that
should not have differences and then show how an incorrect (but widely
used) statistical procedure will result in convincing evidence for
differences between them.**

**Essentially, we think that the core of the problem is the incorrect
statistical procedure (selecting genes by comparing significance with lack
thereof), which will not become more correct by increasing the sample size.**

@@

The authors use a small log-fold change value of 1 to define the
differential expression. This leads to a high number of false positives.

**log2-fold change cutoffs (on top of FDR control) are widely used in the
literature, and a cutoff of 1 is in our experience not small for human
data. The threshold serves to eliminate genes that show statistically
significant change (they are true positives), but are not biologically
relevant (because the effect size, for which log2 fold change is a proxy,
is too small). In any case, the false discovery rate in differential
expression analysis is controlled via BH by a cutoff on the p-value, not
the log fold change, except indirectly (beause log fold change is
correlated to the p-value, setting a more conservative log2 fold change
results in eliminating most of the larger p-values). The precise effect of
raising the LFC threshold on FDR is therefore unclear except in very broad
terms.**

**We have now included a section in the manuscript which shows the
dependence of the artifacts on the log2 fold change thresholds used. For
this, we have replicated the whole procedure of group randomization 100
times. Summary overview can be found in the new Fig. 4. We find that
increasing the threshold can, of course, result in not observing any
enrichments (see Fig. 4, A and B). Without a sufficient number of DEGs, no
enrichments can be calculated using a hypergeometric test. However, even
though increasing the log2 fold change cutoff may sometimes eliminate the
symptoms, it does not remedy the core problem of erroneous analysis of
interactions. Many replicates in the simulations have artifactual
enrichments even at high log fold change thresholds.**

### 2022-01-18 mk4/manuscript.rmd
forgotten update

---
title: ""
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
outputoff:
  html_document
output:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

@@

Version 1: In statistics, comparing the effect of a treatment in two different
conditions requires the use of an appropriate statistical test to assert
the significance of interaction between the treatment and the condition.
However, a common procedure is to define genes specifically regulated in
one, but not the other condition by selecting genes which show a
statistically significant difference in one condition only. Here we show
that not only is this procedure incorrect, but moreover, when combined with
gene set enrichment analyses, it leads to statistically significant, but
incorrect results. These gene set enrichments are the effect of false
negatives and reflect only the differences in gene expression due to
treatment and not due to the difference between conditions. Thus, the
combination of incorrect statistical analysis and gene set enrichment
analysis may result in particularly misleading artifacts.

Version 2: A common application of differential expression analysis is to
ask which genes are differentially expressed upon treatment in one
condition only. One of the approaches is to define genes specific to a
condition by testing for significant difference in expression separately
in the two conditions, and then selecting genes which show statistical
significance in one condition only. This approach may then be combined with
a gene set enrichment analysis to find pathways and gene sets
characteristic for this condition only.

Here we show that this procedure is incorrect. Instead, the interaction
between treatment and condition should be tested. Moreover, we show that
gene set enrichment analysis applied to incorrectly defined genes specific
to one condition only may result in misleading artifacts. Genes which show
significant difference upon treatment in one, but not the other condition
are enriched in gene sets related to the treatment, but not to the
condition. Thus, while the results are apparently related to the treatment,
they do not reflect differences in expression specific to a given
condition.

Version 3:

@@

```{r setup}
padj_thr <- 0.05
lfc_thr  <- 1
n_stratum <- 10 # 2 strata (male and female) per condition/group
                # total number of samples n_stratum * 2 * 2 * 2
lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
```

@@

Transcriptomic analyses commonly include more than one factor. Often, the
question asked is whether there is a difference between conditions (first
factor) with respect to reaction to a particular treatment (second
condition). For example, we may ask whether there are differentially
expressed genes (DEGs) which are specific to a particular group of
patients, such as interferon response elicited by a virus in one group, but
silenced in the other group of patients. In other words, we ask whether the
difference between control (healthy subjects) and treatment (infected
patients) is different between two groups of individuals. This "difference
of differences" is known in statistics as an interaction
[@blalock1965theory]. To find out whether it is statistically significant,
an appriopriate statistical test for interaction should be employed.

However, another approach is widely spread [@nieuwenhuis2011erroneous].
Instead of testing the interaction, one tests the effect of the treatment
separately in both groups. Next, a difference between conditions is inferred if the difference
is significant in one comparison, but not significant in the other. This
approach is not correct from statistical point of view, as "the difference
between significant and not significant is not itself statistically
significant" [@gelman2006difference]. For example, the p-value in the first
comparison may be 0.009, and in the other comparison 0.011, thus at an
alpha level of 0.01 the difference will be statistically significant in the
first, but not significant in the other comparison.

In transcriptomics, the statistical tests are performed for thousands of
genes. As in the general case, the correct inference of differences between
the conditions is by testing the significance of interaction. However, in
another approach the differences between treatment and control are tested
in the two conditions separately, and the results are then compared. This
can be visualized using a Venn diagram (VD, Fig. 1) showing the overall
number of DEGs significant in both comparisons (the intersection in the VD)
or significant in only one comparison (the remaining two fields on a VD).
The genes which are significant in only one comparison may then be
incorrectly defined as specific to the corresponding condition. Following
this, gene set enrichment analysis may be applied to these genes in an
attempt to test which pathways are specific to one, but not the other
condition.

In this paper, we show that under reasonable assumptions this approach may
result in apparent enrichments even if there are no real differences
between the groups. To this end, we have randomly split a cohort into two
groups, compared the treatment (viral infection) with controls in each of
the groups separately and then applied gene set enrichment to the sets of
genes significantly different in one, but not the other group.
Moreover, we show that the resulting gene set enrichments correspond to the
differential expression between treatment and control. Thus, the enriched
terms are relevant to the studied project, yet while they do reflect real
processes linked to treatment, they are not demonstrative of differences
between the study groups. Finally, we use literature search to show that
this incorrect approach to study group-specific treatment effects is not
uncommon.

@@

First, we have performed differential gene expression analysis for each of
the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
DEGs
by using an false discovery rate (FDR) threshold of 
`r padj_thr`
and absolute log~2~ fold change (LFC) threshold of
`r lfc_thr`. There were
`r sum(res$g1$DEG)` 
DE genes in the G1 group, and
`r sum(res$g2$DEG)`
in the G2 group. In total,
`r sum(res.merged$DEG.g1 & res.merged$DEG.g2)` 
DEGs were common for G1 and G2, 
`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
DEGs were significant in G1 only ("specific" for G1), and 
`r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
were significant in G2 only (see Fig. 1, panel *A*).
A naive interpretation of these results implies that there is a substantial
difference between these two groups of individuals, as evidenced by a small
overlap in commonly regulated genes, while the majority of DEGs is
significant in one comparison only.

@@

**Fig. 1. Results of differential gene expression analysis
and gene set enrichment analysis using an incorrect Venn diagram driven approach.**
**A**, Venn diagram showing numbers of differentially expressed genes (DEG)
in each of the two groups, G1 and G2; **B**, Venn diagram showing numbers
of significantly enriched GO terms in each of the two groups; **C** results
of gene set enrichment analysis for genes "specific" to group G1; **D**, results of gene set
enrichment analysis for genes "specific" to group G2 (only top 10 terms are shown).

@@

The groups G1 and G2 were randomly sampled from the same data set.
In fact, repeated re-sampling always results in some genes being found to
be significantly different in one group, but not the other, despite the
fact that one does not expect any major differences between sets of
individuals randomly drawn from one population. Thus, the conclusions drawn
from a Venn diagram-driven gene set enrichment analysis are artefactual. Closer
inspection of genes which belong to DEG in one group, but not the other
reveals the underlying statistical fallacy (Fig. 2, *A* – *D*), that is, that difference
between significant and non-significant is, in itself, not statistically
significant [@gelman2006difference]. This does not necessarily mean that
there are no differences at all between these two groups, but that lack of
significance in one group and significance in the other group does not
correctly identify differences between groups.

@@

To understand how gene set enrichment actually gives significant
results in such randomly generated gene sets, despite absence of genes with
significant interaction, it is first necessary to consider the definition
of a differentially expressed gene in this context. More often than not,
DEGs are defined by a threshold in p-value adjusted for multiple testing,
possibly combined with a threshold in log~2~ fold change. The commonly used
Benjamini-Hochberg procedure [@benjamini1995controlling] ensures that among
genes for which the adjusted p~adj~ < 0.05 there are at most 5% false
positives.

@@

These FNs will occur
at a much higher rate within the sets of DEGs defined by the
non-overlaping areas of the VDs, that is DEGs considered to be "specific"
for one group or other in a naive approach.
To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
were drawn (Fig. 3), comparing the 
`r sum(covar$disease == "no") `
healthy controls to 
`r sum(covar$disease == "SC2") `
COVID-19 patients. Of the
`r length(g1_spec)`
genes significant in G1, but not in G2, 
`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n`
(`r sprintf("%.0f%%", .n/length(g1_spec) * 100)`) 
are significant in the full data set; 
of the
`r length(g2_spec)`
genes significant in G2, but not in G1, 
`r .n <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n`
(`r sprintf("%.0f%%", .n/length(g2_spec) * 100)`) 
are significant in the full data set.
Given that G1 and G2 were sampled from the total population, and since the
FDR was set to `r padj_thr`, we can assume that at least between a third and a half of
the genes that appeared to be "specific" in the initial analysis were, in
fact, false negatives in one of the comparisons.

@@

@nieuwenhuis2011erroneous observed incorrect analyses of interactions in
about half of the analysed papers from top neuroscience journals where the
authors considered an interaction. We wanted to know if this problem is
common in transcriptomics, too. To this end, we have searched three
journals – from broad to specialized – for the occurence of the terms
"differential expression" with "venn diagrams" (Tab. 2). Next, we analysed
the selected papers from year 2020 (and years 2015-2020 in one case) to
decide whether the VD was described or referred to as showing genes
"specific" or "unique" to a particular condition or whether a test for
interaction was performed. Finally, we checked whether gene set enrichment
analysis was applied to genes significant in one, but not the other
condition in order to find gene sets and pathways regulated by one, but not
the other condition.

@@

Our literature survey showed that even in cases where a VD is absent from a
paper, at least two other ways are not uncommon. Firstly, the direct
comparison of gene set enrichment results: that is, drawing conclusions
from the fact that a gene set enrichment result was significant in one
comparison only.  Second, while Venn diagrams are often used to illustrate
the numbers of "specific" DEGs and so present a mean to find examples of
this fallacy in scientific literature, researches often test for enrichment
these "specific" genes without using the phrase "Venn diagram" or even
clearly stating how the lists of "specific" genes were derived. In all
these cases, the analysis boils down to comparing results significant in
one, but not in another comparison.

@@

Defining condition-specific genes based on significant difference in one,
but no significant difference in another comparison is thus not only a
statistical fallacy, but when combined with gene set enrichment analysis
also one which can lead to potentially sound-looking results. This method
of obtaining specific differences between conditions should therefore be
abandoned in favor of statistically correct approaches.
Furthermore, gene set enrichment analysis must never be applied
to sets of genes defined as significant in one comparison, but not the
other.

@@

@@

A literature survey was performed using Google Scholar to
estimate the frequency of the incorrect use of Venn diagrams. We searched
for articles from 2020 mentioning the phrases "differential expression" and "venn
diagram" in the Scientific Reports journal. The journal was
selected because it represents a wide spectrum of research areas, the
articles are available through an open access license and there is a large
number of publications per year. For each of the papers identified, we
checked whether (i) the authors used the VD to show differentially expressed
transcripts significant in one comparison, but not another, (ii) the
authors discussed "unique", "non-overlapping" or "specific" regions of the
Venn diagram and (iii) whether this was coupled to gene set enrichment
analysis is any form. Articles which (i) focused only on the intersections
of the Venn diagrams (genes common to all conditions), or (ii) which used the
Venn diagrams for a purpose other than to compare genes significant in one
condition, but not significant in another condition or (iii) for which a
clear-cut error could not be indicated past any reasonable doubt were not
considered incorrect.
The results of the literature survey (including links to papers classified
as incorrectly analysing the interaction) are included in the manuscript
sources.

>>>>

---
title: "Venn diagrams as indicators of erroneous statistical reasoning in transcriptomics."
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
outputoff:
  html_document
output:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

@@

A common application of differential expression analysis is to
find genes that are differentially expressed upon treatment in only one
out of several groups of samples. One of the approaches is to test 
for significant difference in expression between treatment and control separately
in the two groups, and then select genes that show statistical
significance in one group only. This approach is then often combined with
a gene set enrichment analysis to find pathways and gene sets
regulated by treatment in only this group.

Here we show that this procedure is statistically incorrect and that the interaction
between treatment and group should be tested instead. Moreover, we show
that gene set enrichment analysis applied to such incorrectly defined genes
group-specific genes only may result in misleading artifacts. Genes which
show significant difference upon treatment in one, but not another group
are enriched in gene sets associated with the treatment in general.  Thus,
the results appear related to the problem at hand, but do not reflect the
group-specific effect of a treatment. A literature search revealed that
more than a quarter of papers which used a Venn diagram to illustrate the
results of separate differential analysis  have also applied this incorrect
reasoning.

@@

```{r setup}
padj_thr <- 0.05
lfc_thr  <- 1
n_stratum <- 10 # 2 strata (male and female) per group/treatment
                # total number of samples = n_stratum * 2 * 2 * 2
lfc_thresholds <- c(0, 0.5, 1, 1.5, 2, 2.5, 3)
```

@@

Experimental designs for transcriptomic analyses commonly include more than one factor. Often, the
question asked is whether there is a difference between groups (first
factor) with respect to reaction to a particular treatment (second
factor). For example, we may ask whether there are differentially
expressed genes (DEGs) which are specific to a particular group of
patients, e.g. interferon response elicited by a virus in one group, but
not regulated in another group of patients. In other words, we ask whether the
difference between control (healthy subjects) and treatment (infected
patients) is different between two groups of individuals. This "difference
of differences" is known in statistics as an interaction
[@blalock1965theory]. To find out whether it is statistically significant,
an appriopriate statistical test for interaction should be employed.

However, another approach is widely spread [@nieuwenhuis2011erroneous].
Instead of testing the interaction, the effect of the treatment is tested
separately in both groups. Next, a difference between groups is inferred if the difference
is significant in one comparison, but not significant in the other. This
approach is not correct from statistical point of view, as "the difference
between significant and not significant is not itself statistically
significant" [@gelman2006difference]. For example, the p-value in the first
comparison may be 0.009, and in the other comparison 0.011, thus at an
alpha level of 0.01 the difference will be statistically significant in the
first, but not significant in the other comparison.

In transcriptomics, statistical tests are performed for thousands of
genes. As in the general case, the inference of differences between the
groups should correctly be done by testing the significance of interaction
between the group and the treatment. However, often the differences between
treatment and control are tested in the two groups separately, and the
results are then compared. This can be visualized using a Venn diagram (VD,
Fig. 1) showing the overall number of DEGs significant in both comparisons
(the intersection in the VD) or significant in only one comparison (the
remaining two fields on a VD).  The genes which are significant in only one
comparison may then be incorrectly defined as specifically to the corresponding
group. Following this, gene set enrichment analysis may be applied to
these genes in an attempt to test which pathways are specific to one, but
not the other group.

In this paper, we show that under reasonable assumptions this approach may
result in apparent enrichments even if there are no real differences
between the groups. To this end, we randomly split a cohort into two
groups, compared the treatment (viral infection) with controls in each of
the groups separately and then applied gene set enrichment to the sets of
genes significantly different in one, but not the other group.
Moreover, we show that the resulting gene set enrichments correspond to the
differential expression between treatment and control. Thus, the enriched
terms are relevant to the studied project, yet while they do reflect real
processes linked to treatment, they are not demonstrative of differences
between the study groups. Finally, we use literature search to show that
this incorrect approach to study group-specific treatment effects is not
uncommon. In fact, while VDs are a useful visualisation tool also in
transcriptomics, in more than quarter of the papers where VDs were used,
group-specific genes were defined as significant in one, but not other
groups, and in 19% a gene set enrichment was performed.

@@

First, we have performed differential gene expression analysis for each of
the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
DEGs
by using an false discovery rate (FDR) threshold of 
`r padj_thr`
and absolute log~2~ fold change (LFC) threshold of
`r lfc_thr`. There were
`r sum(res$g1$DEG)` 
DEGs in the G1 group, and
`r sum(res$g2$DEG)`
in the G2 group. In total,
`r sum(res.merged$DEG.g1 & res.merged$DEG.g2)` 
DEGs were common for G1 and G2, 
`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
DEGs were significant in G1 only ("specific" for G1), and 
`r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
were significant in G2 only (see Fig. 1, panel *A*).
A naive interpretation of these results implies that there is a substantial
difference between these two groups of individuals, as evidenced by a small
overlap in commonly regulated genes, while the majority of DEGs is
significant in one comparison only.

@@

**Fig. 1. Results of differential gene expression analysis
and gene set enrichment analysis using an incorrect approach.**
**A**, Venn diagram showing numbers of differentially expressed genes (DEG)
in each of the two groups, G1 and G2; **B**, Venn diagram showing numbers
of significantly enriched GO terms in each of the two groups; **C** results
of gene set enrichment analysis for genes "specific" to group G1; **D**, results of gene set
enrichment analysis for genes "specific" to group G2 (only top 10 terms are shown).

@@

The groups G1 and G2 were randomly sampled from the same data set.
In fact, repeated re-sampling always results in some genes being found to
be significantly different in one group, but not the other, despite the
fact that one does not expect any major differences between sets of
individuals randomly drawn from one population. Thus, the conclusions drawn
from a Venn diagram-driven gene set enrichment analysis are artefactual. Closer
inspection of genes which are DEGs in one group, but not the other
reveals the underlying statistical fallacy (Fig. 2, *A* – *D*), that is, that difference
between significant and non-significant is, in itself, not statistically
significant [@gelman2006difference]. This does not necessarily mean that
there are no differences at all between these two groups, but that lack of
significance in one group and significance in the other group does not
correctly identify differences between groups.

@@

To understand how gene set enrichment actually gives significant
results in such randomly generated gene sets, despite absence of genes with
significant interaction, it is first necessary to consider the definition
of a differentially expressed gene in this context. More often than not,
DEGs are defined by a threshold in p-value adjusted for multiple testing,
possibly combined with a threshold in log~2~ fold change. The commonly used
Benjamini-Hochberg procedure [@benjamini1995controlling] ensures that among
genes for which the adjusted p~adj~ < 0.05 there are at most 5% false
positives irrespective of the sample size.

@@

These FNs will occur
at a much higher rate within the sets of DEGs defined by the
non-overlaping areas of the VDs, that is DEGs considered to be "specific"
for one group or other in a naive approach.
To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
were drawn (Fig. 3), comparing the 
`r sum(covar$disease == "no") `
healthy controls to 
`r sum(covar$disease == "SC2") `
COVID-19 patients. Of the
`r length(g1_spec)`
genes significant in G1, but not in G2, 
`r .n1 <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n1`
(`r sprintf("%.0f%%", .n1/length(g1_spec) * 100)`) 
are significant in the full data set; 
of the
`r length(g2_spec)`
genes significant in G2, but not in G1, 
`r .n2 <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n2`
(`r sprintf("%.0f%%", .n2/length(g2_spec) * 100)`) 
are significant in the full data set.
Given that G1 and G2 were sampled from the total population, and since the
FDR was set to `r padj_thr`, we do not expect more than 
`r .nfp <- ceiling(padj_thr * sum(res$full$DEG)); .nfp`
FPs in the full data set, 
which implicates that at least 
`r .n1 + .n2 - .nfp` out of the `r length(g1_spec) + length(g2_spec)` 
"specific" DEGs are true positives in the full data set.
Thus, we can assume that at least a third of
the genes that appeared to be "specific" in the initial analysis were, in
fact, false negatives in one of the comparisons.

@@

@nieuwenhuis2011erroneous observed incorrect analyses of interactions in
about half of the analysed papers from top neuroscience journals where the
authors considered an experimental design allowing for a test for
interaction. We wanted to know if this problem is
common in transcriptomics, too. To this end, we have searched three
journals – from broad to specialized – for the occurence of the terms
"differential expression" with "venn diagrams" (Tab. 2). Next, we analysed
the selected papers from year 2020 (and years 2015-2020 in one case) to
decide whether the VD was described or referred to as showing genes
"specific" or "unique" to a particular group or whether a test for
interaction was performed. Finally, we checked whether gene set enrichment
analysis was applied to genes significant in one, but not the other
group in order to find gene sets and pathways regulated by one, but not
the other group.

@@

While VDs appear to be frequently associated with an incorrect statistical
reasoning, the use of VDs is not the cause.  In the absence of a VD
illustrating the DEGs common and unique to the different study groups, two
incorrect approaches may still be found. Firstly, the direct
comparison of gene set enrichment results: that is, drawing conclusions
from the fact that a gene set enrichment result was significant in one
comparison only.  Second, while VDs are often used to illustrate
the numbers of "specific" DEGs and so present a mean to find examples of
this fallacy in scientific literature, researches test for enrichment
these "specific" genes without using the phrase "Venn diagram" or even
clearly stating how the lists of "specific" genes were derived. In all
these cases, the analysis boils down to comparing results significant in
one, but not in another comparison.

@@

Defining group-specific genes based on significant difference in one,
but no significant difference in another comparison is thus not only a
statistical fallacy, but when combined with gene set enrichment analysis
also one which can lead to potentially sound-looking results. This method
of obtaining specific differences between groups should therefore be
abandoned in favor of statistically correct approaches.
Furthermore, gene set enrichment analysis must never be applied
to sets of genes defined as significant in one comparison, but not the
other.

@@

### Simulation study

We have generated replicates of the example study for different log~2~ fold
change thresholds (`r paste(lfc_thresholds, sep=", ")`) and three different
sample sizes (40, 80 and 160, corresponding to sample size per
group/treatment combination of 10, 20 and 40). For each replicate, the full
procedure as described above was repeated, and numbers of DEGs and
significantly enriched terms were collected.

@@

A literature survey was performed using Google Scholar to
estimate the frequency of the incorrect use of Venn diagrams. We searched
for articles from 2020 mentioning the phrases "differential expression" and "venn
diagram" in three journals: Scientific Reports (2020), Nature Communications (2020) and
Science Immunology (2015-2020). For each of the papers identified, we
checked whether (i) the authors used the VD to show differentially expressed
transcripts significant in one comparison, but not another, (ii) the
authors discussed "unique", "non-overlapping" or "specific" regions of the
Venn diagram and (iii) whether this was coupled to gene set enrichment
analysis is any form. Articles which (i) focused only on the intersections
of the Venn diagrams (genes common to all groups), or (ii) which used the
Venn diagrams for a purpose other than to compare genes significant in one
groups, but not significant in other groups or (iii) for which a
clear-cut error could not be indicated past any reasonable doubt were not
considered incorrect.
The results of the literature survey (including links to papers classified
as incorrectly analysing the interaction) are included in the manuscript
sources.
